---
title: "Metabolomics Data Matrix"
author: "Pietro Franceschi"
date: "12/2/2020"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
```

## Many variables in many samples

As in all _omic_ experiments, a metabolomic assay allows to measure a large number of _variables_(properties) in a (hopefully) reasonably large number of samples.

<hr>

**... what are the varibles ...**

> 1. In a targeted metabolomics investigation
> 2. In an untargeted MS based metabolomics analysis
> 3. In an untargeted NMR based metabolomics assay

	

## Number of variables ... place your guess

> 1. from 10 to 300
> 2. from 1000 to 30000
> 3. from 500 to 1000


## Data Matrix

```{r, echo=FALSE, fig.align='center', out.width="60%"}
include_graphics("images/Matrix.svg")
```

## Variables (columns) are not independent

* Biological relations
* Chemical/Analytical reasons

```{r fig.align='center'}
x <- 1:20
y <- x*2 + rnorm(20,0,3)
par(pty="s")
plot(x,y, main = "Correlation", xlab = "Var1", ylab = "Var2", cex = 1, col = "coral", pch = 19)
```

## Rows (samples) are often dependent

The design of the study can result in a multilevel hierarchical structure of the samples which violate independence. In other words some samples are "by construction" associated and share something in common

**Examples**

1. Repeated measures of the same individual
2. Subpopulations
3. Different site, different day, ...

## Data Matrix size

In the typical "happy" statistical context, the number of variable we measure is smaller than the number of samples. 

In _omics_ the number of variables normally outperforms the number of samples

<hr>

<div class="red2">
*FAT DATA MATRIX*
<div>


```{r fig.align='center', fig.height=2.5, fig.width=10, out.width="100%"}
mymat <- matrix(rnorm(20*300), nrow = 20)
par(mar=rep(0, 4), xpd = NA) 
image(t(mymat), asp = 20/300, xaxt='n', ann=FALSE, yaxt='n', bty="n")
```

## Univariate approach

The **Univariate** appraoch considers each variable separately and it applies "standard" statistical tools to spot the more **interesting** variables

1. Statistical testing 
2. Linear modeling (`lm`, `glm`, ...)
3. ANOVA
4. Hierarchical Modeling

```{r fig.align='center', fig.height=2.5, fig.width=6, out.width="60%"}
mymat <- matrix(rnorm(20*100), nrow = 20)
par(mar=rep(0, 4), xpd = NA) 
image(t(mymat), asp = 20/100, xaxt='n', ann=FALSE, yaxt='n', bty="n")
rect(xleft = 0.5, xright = 0.51, ybottom = -0.1, ytop = 1.1, border = "steelblue", lwd = 2)
rect(xleft = 0.3, xright = 0.31, ybottom = -0.1, ytop = 1.1, border = "steelblue", lwd = 2)
rect(xleft = 0.7, xright = 0.71, ybottom = -0.1, ytop = 1.1, border = "steelblue", lwd = 2)
```

## Multivariate Approach

Each samples is represented as a point in the multidimesional variable space. The dataset is a **cloud** of points in that space.

The size of the space equals the number of variables we are measuring. **Multivariate methods** (PCA, PLS, ASCA, ...) are able to exploit the structure of the space

```{r, echo=FALSE, fig.align='center', out.width="40%"}
include_graphics("images/Coord_planes_color.svg")
```

## Why multivariate

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(cowplot)
library(ggpubr)

ggscatterhist(
  iris, x = "Sepal.Length", y = "Sepal.Width",
  color = "Species", # comment out this and last line to remove the split by species
  #margin.plot = "histogram", # I'd suggest removing this line to get density plots
  margin.params = list(fill = "Species", color = "black", size = 0.2)
) 

```

## Multivariate


**PRO**

1. Potentially more powerful
2. Explicit use of variable correlation
3. No issues with multiple testing

<hr>

**CONS**


1. Chance correlations in fat matrices
2. Empty Space
3. Difficult to embed hierarchical structure

## Univariate


**PRO**


1. Statistical modeling is there!
2. Interpretable by construction

<hr>

**CONS**


1. Multiple testing
2. The structure of the data creates redundancy


## The course of dimensionality

```{r fig.align='center', fig.height=2.5, fig.width=8, out.width="100%"}
set.seed(123)

par(mfrow = c(1,3))
par(pty="s")
plot(y = rep(0,10), x = runif(10,0,10), xaxt='n', yaxt='n', bty="n", xlim = c(-1,11), 
     main = "10 points 1D", xlab = "", ylab = "", col = "steelblue", cex = 1.5, pch  =21)
arrows(x0 = 0 , y0 = 0, x1 = 11)

par(pty="s")
plot(y = runif(10,0,10), x = runif(10,0,10), xlab = "D1", ylab = "D2", main = "10 points 2D", col = "steelblue", cex = 1.5, pch  =21)


plot(y = runif(100,0,10), x = runif(100,0,10), xlab = "D1", ylab = "D2", main = "100 points 2D", col = "steelblue", cex = 1.5, pch  =21)

```

1. To fill the space the number of points is not linear with the number of dimensions
2. Already with 10 samples the 2d plot looks empty
3. Can you imagine 20 samples in 10000 dimensions? ;-)

 
## Practical !

```{r, echo=FALSE, fig.align='center', out.width="50%"}
include_graphics("images/drill.svg")
```





