---
title: "Multiple Testing"
author: "Pietro Franceschi<br>pietro.franceschi@fmach.it"
output:
  revealjs::revealjs_presentation:
    theme: white
    highlight: pygments
    css: "../mycss.css"
    transition: slide
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(knitr)
library(tidyverse)
```

## Where we left off

In a single test, we accept a ~~false positive rate~~ of $\alpha = 0.05$.

. . .

This means: if H0 is true, we wrongly reject it **5% of the time**.

. . .

<div class="box-red">
⚠️ <strong>The metabolomics reality:</strong><br>
We do not test <em>one</em> hypothesis. We simultaneously test <strong>hundreds or thousands</strong> of variables.
</div>

## A concrete example

Imagine a metabolomics experiment:

::: incremental
- **2 groups** of samples (treated vs control)
- **1000 metabolites** measured for each sample
- A **t-test** for every metabolite → 1000 p-values
- We flag anything with $p < 0.05$ as "significant"
:::

. . .

How many false positives should we expect?

## Expected false positives

If ~~none~~ of the 1000 metabolites truly differ between groups:

$$\text{Expected false positives} = 1000 \times 0.05 = \mathbf{50}$$

. . .

We would report **50 "significant" metabolites** — all of them artefacts.

. . .

<div class="box-key">
Performing many tests inflates the probability of finding something by chance alone.
</div>

# Visualising the problem

## p-values under H₀

When there is **no real effect**, p-values follow a ~~uniform distribution~~ between 0 and 1.

```{r fig.align='center', out.width="50%"}
set.seed(42)
n_vars <- 1000
n_samples <- 20

A <- matrix(rnorm(n_vars * n_samples), nrow = n_samples)
B <- matrix(rnorm(n_vars * n_samples), nrow = n_samples)

pvals <- sapply(1:n_vars, function(i) t.test(A[, i], B[, i])$p.value)

tibble(p = pvals) %>%
  ggplot(aes(x = p)) +
  geom_histogram(binwidth = 0.05, fill = "steelblue", colour = "white") +
  geom_vline(xintercept = 0.05, colour = "#e41a1c", linetype = "dashed", linewidth = 1) +
  annotate("text", x = 0.08, y = 65, label = "α = 0.05", colour = "#e41a1c", hjust = 0) +
  labs(x = "p-value", y = "Count",
       title = paste0("1000 random t-tests — ", sum(pvals < 0.05), " with p < 0.05")) +
  theme_light(base_size = 16)
```

## What this tells us

::: incremental
- The histogram is **flat** — every p-value is equally likely under H0
- The red bar on the left ($p < 0.05$) contains ~50 tests
- These are ~~all false positives~~ — there is no true signal in this data
- A flat p-value histogram is actually a useful **quality check**: it confirms no real effects are present
:::

## When there ARE real effects

If some variables truly differ between groups, the histogram is ~~no longer flat~~: a spike appears near zero on top of the uniform background.

```{r fig.align='center', out.width="50%"}
set.seed(99)
n_null_h   <- 950   # variables with no effect
n_signal_h <- 50    # variables with a true effect
n_samp_h   <- 20

pv_null   <- sapply(1:n_null_h, function(i)
  t.test(rnorm(n_samp_h), rnorm(n_samp_h))$p.value)

pv_signal <- sapply(1:n_signal_h, function(i)
  t.test(rnorm(n_samp_h, mean = 0), rnorm(n_samp_h, mean = 1.8))$p.value)

tibble(p     = c(pv_null, pv_signal),
       origin = c(rep("null (no effect)", n_null_h),
                  rep("true effect", n_signal_h))) %>%
  ggplot(aes(x = p, fill = origin)) +
  geom_histogram(binwidth = 0.05, colour = "white", position = "stack") +
  geom_vline(xintercept = 0.05, colour = "#e41a1c", linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("null (no effect)" = "grey70", "true effect" = "#377eb8"),
                    name = "") +
  labs(x = "p-value", y = "Count",
       title = "1000 tests: 950 null + 50 true effects") +
  theme_light(base_size = 16) +
  theme(legend.position = "bottom")
```

## Reading the mixed histogram

::: incremental
- The **flat part** across the full range → null features (false positives hiding here)
- The **spike near zero** → true positives piling up at small p-values
- The height of the spike above the flat baseline is proportional to the ~~number of real effects~~
- A spike + flat pattern is the signature of a real experiment with **some** true signal
:::


. . .


<br>

<div class="box-blue">
<strong>Practical tip:</strong> always plot the p-value histogram before applying any correction — it tells you immediately whether your data contains any signal at all.
</div>

## Family-wise error rate (FWER)

The probability of making **at least one** false positive across $m$ independent tests:

$$\text{FWER} = 1 - (1 - \alpha)^{m}$$

. . .

| Tests ($m$) | FWER at $\alpha = 0.05$ |
|:-----------:|:-----------------------:|
| 1           | 0.05                    |
| 10          | 0.40                    |
| 50          | 0.92                    |
| 100         | 0.994                   |
| 1000        | ≈ 1                     |

. . .

<br>

<div class="box-red">
⚠️ With the number of tests typical of metabolomics, an uncorrected analysis is <em>guaranteed</em> to produce false positives.
</div>

# Correction methods

## Two families of solutions

::: incremental
1. **Control FWER** — ensure the probability of *any* false positive stays below $\alpha$
   - Bonferroni, Holm
   - Very conservative; appropriate when ~~every false positive is dangerous~~

2. **Control FDR** — ensure the *proportion* of false positives among findings stays below $\alpha$
   - Benjamini-Hochberg
   - Less conservative; appropriate for ~~discovery science~~
:::

## Bonferroni correction

The simplest FWER control: divide the threshold by the number of tests.

$$\alpha_{\text{Bonf}} = \frac{\alpha}{m}$$

. . .

For $m = 1000$ and $\alpha = 0.05$:

$$\alpha_{\text{Bonf}} = \frac{0.05}{1000} = 0.00005$$

. . .

<div class="box-red">
⚠️ <strong>Very conservative.</strong> A true positive must produce an extremely small p-value to survive. In metabolomics with correlated variables this is often too strict.
</div>

## False Discovery Rate (FDR)

Introduced by **Benjamini & Hochberg (1995)** — one of the most cited papers in statistics.

. . .

Instead of controlling the probability of *any* false positive, control the **expected proportion** of false positives among all discoveries:

$$\text{FDR} = E\left[\frac{\text{false positives}}{\text{total discoveries}}\right]$$

. . .

The resulting corrected p-values are called ~~q-values~~.

A q-value of 0.05 means: among all features with $q \leq 0.05$, we expect at most **5% to be false positives**.



# The volcano plot

## Combining p-value and effect size

A **volcano plot** displays, for each variable:

- x-axis: ~~fold change~~ (or log₂ fold change) — the *size* of the effect
- y-axis: $-\log_{10}(p)$ — the *confidence* in the effect

```{r fig.align='center', out.width="50%"}
set.seed(123)
n_feat <- 500

fc  <- rnorm(n_feat, 0, 0.8)
pv  <- runif(n_feat)

# inject some real signal
sig_idx <- sample(1:n_feat, 40)
fc[sig_idx]  <- rnorm(40, 0, 1) + sample(c(-2.5, 2.5), 40, replace = TRUE)
pv[sig_idx]  <- runif(40, 0, 0.001)

qv <- p.adjust(pv, method = "BH")

tibble(fc = fc, pv = pv, qv = qv) %>%
  mutate(status = case_when(
    qv < 0.05 & fc >  1 ~ "up",
    qv < 0.05 & fc < -1 ~ "down",
    TRUE                 ~ "ns"
  )) %>%
  ggplot(aes(x = fc, y = -log10(pv), colour = status)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_hline(yintercept = -log10(max(pv[qv < 0.05])),
             linetype = "dashed", colour = "grey40") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", colour = "grey40") +
  scale_colour_manual(values = c(up = "#e41a1c", down = "#377eb8", ns = "grey60"),
                      name = "") +
  labs(x = "log₂ fold change", y = "-log₁₀(p)") +
  theme_light(base_size = 16) +
  theme(legend.position = "bottom")
```

## Reading a volcano plot

::: incremental
- Points in the **upper corners** — large effect, high confidence → strong candidates
- Points at the **top centre** — significant but small effect → statistically real, biologically questionable
- Points on the **sides, low** — large fold change but high p-value → likely noise or low sample size
- The horizontal dashed line marks the **FDR threshold**
- The vertical lines mark the **fold-change cutoff**
:::

# Take-home messages

##

<div class="box-red">
⚠️ <strong>Never report uncorrected p-values</strong> when testing multiple hypotheses simultaneously.
</div>

<br>

::: incremental
- The **number of tests** is as important as the threshold
- **Bonferroni** for strict FWER control (clinical, confirmatory settings)
- **Benjamini-Hochberg** for discovery (metabolomics, genomics, exploratory)
- Always inspect the **p-value histogram** — a spike near 0 and flat elsewhere is a good sign
- Statistical significance alone is never enough: pair it with ~~effect size~~
:::

##

<div class="box-key">
In a typical metabolomics experiment you will test hundreds of features. Without correction, your results list is dominated by noise. With FDR control, you can make principled discoveries.
</div>
