---
title: "Stratified Random Sampling"
author: "Pietro Franceschi"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, error = FALSE)
library(tidyverse)
```


## Intro

During the presentation I clearly affirmed that in presence of subpopulations **stratified random sampling** is able to give a more accurate estimate of the properties of my reference population. Let's check my statement ...


## The question
Suppose that I'm interested in estimating the average height of group of 10000 people (6000 males and 4000 females). The two subpopulations are characterized by the following properties, which normally are unknown ...

* Males: mean = 180 cm, sd = 10 cm
* Females: mean = 150 cm, sd = 10 cm

## Running the simulations

First of all I create my population

```{r}
## separate tibbles
males <- tibble(gender = "M", height = rnorm(8000,180,10))
females <- tibble(gender = "F", height = rnorm(2000,150,10))

## merge everything in a df
mypop <- males %>% 
  bind_rows(females)
```


Just for fun let's give a look 

```{r}
mypop %>% 
  ggplot() + 
  geom_histogram(aes(x = height, fill = gender), col = "white",  alpha = 0.6, position = "identity") + 
  theme_light()
```

Ok, we see what we expect, the two groups are different, and there is an unbalancing of the two subpopulations.




## The problem

suppose that I'm interested in estimating well the true population mean. In my case this parameter is known ...

```{r}
pop_mu <- mean(mypop$height)
pop_mu
```

```{r}
mypop %>% 
  ggplot() + 
  geom_histogram(aes(x = height, fill = gender), col = "white",  alpha = 0.6, position = "identity") + 
  geom_vline(xintercept = pop_mu, col="red", lty = 2) + 
  theme_light()
```

What I would do is to **sample** a group of people from the population and use the mean of this group as an estimate of the population mean. The number of samples I can extract is normally dictated by practical or economical reasons.


## Simple Random Sampling
suppose that I have no clue about the presence of the two sub-populations, so the best I can do to avoid biases is **simple random sampling**.

Suppose that I can afford to take 30 people and we use this group to estimate the population mean, what is the type of variability we expect with a simple **Random Sampling**?

To do that I simulate the extraction of 30 people from my population and to see the variability of the process I do it 500 times

```{r}
## start creating the tibble container
x <- tibble(sampid = paste0("s",1:500))
head(x)
```

```{r}
## here we do the complete random sampling ...
x <- x %>% 
  mutate(random_samples = map(sampid, ~sample(x = mypop$height,size = 30)))
head(x)
```


```{r}
## now we calculate the averages
x <- x %>% 
  mutate(random_averages = map_dbl(random_samples, ~mean(.x)))
head(x)
```

This plot show the variability of the estimate as a reference with the true population mean

```{r}
x %>% 
  ggplot() + 
  geom_point(aes(x = sampid, y = random_averages)) + 
  geom_hline(yintercept = pop_mu, col = "red", lty = 2) + 
  theme_light() + 
  theme(axis.text.x = element_blank())
```


```{r}
x %>% 
  ggplot() + 
  geom_histogram(aes(x = random_averages), alpha = 0.5, col = "white") + 
  geom_vline(xintercept = pop_mu, col = "red", lty = 2) + 
  theme_light()
```


As we can see, the estimate has variance (:-P), and it is centered around the correct value for the population mean. It is also normally distributed, but this is something we will touch later today.

## Stratified sampling

Suppose now that I'm aware of the fact that my population is composed by different subgroups. To take into account that I switch to **stratified random sampling**


If I'm sticking with 30 people I have to allocate them to the two groups in a proportion that mirrors the proportions of the individuals in the population

```{r}
nsamp <- 30

## here I calculate the fraction
subfractions <- (table(mypop$gender)/nrow(mypop))*nsamp %>% ceiling(.)



## just to show
library(magrittr)

another_subfraction <- mypop$gender %>% 
  table(.) %>% 
  divide_by(nrow(mypop)) %>% 
  multiply_by(nsamp) %>% 
  ceiling(.)

subfractions
```


```{r}
## here we do the complete random sampling of the two populations
## and we return a vector with the samples extracted in a stratified way
x <- x %>% 
  mutate(stratified_random_samples = map(sampid, function(n) {
    out_males <- sample(mypop %>% filter(gender == "M") %>% pull(height),subfractions["M"])
    out_females <- sample(mypop %>% filter(gender == "F") %>% pull(height),subfractions["F"])
    return(c(out_males,out_females))
  }))
head(x)
```

Now, the average of the samples will be another time the estimate of the population mean

```{r}
## now we calculate the averages
x <- x %>% 
  mutate(stratified_averages = map_dbl(stratified_random_samples, ~mean(.x)))
head(x)
```

And now we plot!

```{r}
x %>% 
  ggplot() + 
  geom_point(aes(x = sampid, y = random_averages)) + 
  geom_point(aes(x = sampid, y = stratified_averages), col = "red") + 
  geom_hline(yintercept = pop_mu, col = "red", lty = 2) + 
  theme_light() + 
  theme(axis.text.x = element_blank())
```

```{r}
x %>% 
  ggplot() + 
  geom_histogram(aes(x = random_averages), fill = "blue", alpha = 0.5, col = "white") + 
  geom_histogram(aes(x = stratified_averages), fill = "red", alpha = 0.5, col = "white") + 
  geom_vline(xintercept = pop_mu, col = "red", lty = 2) + 
  theme_light()
```


As you can see the average is still good, but the _variability (variance) is much smaller_
This is true for all the tings you would like to estimate from your sample. If you think to it, the reason for this fact is that using stratification we ensure that our sample is representative of both subpopulations.

**Something for You**

* Could you check what happens to the two types of sampling if you change the unbalancing in the population? 
* Could you check what happens to the two types of sampling if you change the size of the sample (here 30)?





