---
title: "Distribution of Measured Variables"
author: "Pietro Franceschi"
output: 
  ioslides_presentation:
    css: "../mycss.css"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(tidyverse)
```



## Normal Distribution {#notes .smaller}

- The Normal (Gaussian) distribution has a prominent role in statistics
- Some sort of "normality" is often the prerequisite of many statistical tools
- With normal data, ~~the mean is the value with higher likelihood~~
- Reasoning on the mean is equivalent is reasoning on the most probable value ...

```{r out.width= "50%", fig.align='center'}
## Standard normal distribution:

xvalues <- data.frame(x = c(-10, 10))

## add a second normal curve
normal1 <- function(x){
  dnorm(x, mean = 0, sd = 3)
}


xvalues %>% ggplot(aes(x = x)) + 
  stat_function(fun = normal1, lwd = 1, col = "darkblue") + 
  stat_function(fun = normal1, geom = "area", fill = "steelblue", alpha = 0.5) +
  theme_light() + 
  geom_hline(yintercept = 0, lwd = 1) + 
  geom_vline(xintercept = 0, lty = 2, col = "red") + 
  geom_text(x = 0, y = 0.1, label = "mean", size =7) + 
  ylab("Likelihood")


```

## Sampling Distribution

The distribution of the mean of a ~~sample~~ extracted from every type of distribution is always normal

```{r fig.width=7, fig.align='center'}
popunif <- runif(10000)

means <- rep(0,500)

for(i in 1:500){
  means[i] <- mean(sample(popunif,10))
}

par(mfrow = c(1,2))
hist(popunif, main = "Arbitrary Population", col = "steelblue")
hist(means, main = "Means of repeated sampling (20)", col = "Orange")

```

## Sampling Distribution {#highlights}

**Noteworthy**

> 1. The sampling distribution get's narrower as the sample increases 
> 2. This is the ultime reasons why measuring more samples leads to smaller _p-values_


## Formally

- Mean of the sampling distribution: $\mu_{pop}$  
- Variance of the sampling distribution: $\frac{\sigma}{\sqrt(N)}$  


Where $N$ is the size of the sample


## ... but {.build}

We are often dealing with data non normally distributed

>1. Count based technologies never return negative numbers ... (MS!)
>2. In many biochemical reactions the product of concentrations is important  
>3. Presence of sub-populations
>4. Outliers

Sub populations are typically present in complex experiments

## Subpopulations

```{r fig.align='center'}
subpop <- c(rnorm(100,10,3),rnorm(100,20,3))

hist(subpop, breaks = 30)
abline(v = c(10,20), col = "red", lty = 2, lwd = 2, main = "2 sub-populations")
```

**NB: In many cases subpopulations are unknown!


## Lognormal data {.smaller}

Lognormal data are extremely common in metabolomics ...

```{r out.width= "70%", fig.align='center'}
set.seed(124)
lndata <- rlnorm(200, meanlog = 0, sdlog = 1)

hist(lndata, breaks = 50, col = "steelblue", main = "Lognormal Data")
abline(v = mean(lndata), col  ="red", lwd = 2)
abline(v = median(lndata), col  ="darkgreen", lwd = 2)
legend("topright", legend = c("mean","median"), col = c("red","darkgreen"), lwd = 1, box.lwd = 0)
```

- The mean is **not** the most probable value!
- Statistical machinery focusing on **mean** is not the right tool for the trade

## Solutions

> 1. **Non-parametric** approaches (Kruskall-Wallis, quantile-regression, permutations, bootstrap, ...)
> 2. Variable **transformations** 


Remember that non parametric tests suffer of lack of power and are often completely useless fin investigations with only few samples (5-10)

---


```{r fig.align='center'}
lndata_t <- log10(lndata)

hist(lndata_t, breaks = 50, col = "steelblue", main = "Log trasformed Lognormal Data")
abline(v = mean(lndata_t), col  ="red", lwd = 2)
abline(v = median(lndata_t), col  ="darkgreen", lwd = 2)
legend("topright", legend = c("mean","median"), col = c("red","darkgreen"), lwd = 1, box.lwd = 0)
```

~~Mean and median are ok!~~


## Checking Normality

1. Normality tests ... again statistics ;-)
2. Graphical methods `q-q plots`

## Quantile - quantile plots

Quantile quantile plots are used to visually compare the **theoretical quantiles** of a distribution with the **sample quantiles** 

```{r fig.align='center'}
a <- rnorm(30)
par(pty="s")
qqnorm(a, main = "Normal q-q plot of normal data", pch = 19)
qqline(a)
```

## q-q plot for lognormal data

```{r fig.align='center'}
par(pty="s")
qqnorm(lndata, main = "Normal q-q plot of log-normal data", pch = 19, col = "#984ea380")
qqline(lndata)

```
 

## Take home message {#highlights}

Before running your statistical machinery give a look to your data to check if what you are doing makes sense ...